Signatures and Induction Principles for Higher Inductive-Inductive Types

### 2019-08-15 response.txt

Response to the first reviewer:

- Extending to n-Type:
  As we note in section 2, generalizing external types in signatures to Type-i
  is possible and not difficult. We omit this generalization from the paper
  mainly in order to sync with the existing Agda formalization.

- Page 1: initiality vs finite generation
  We revised the wording.

- Lambdas in signatures:

  The only function space where it makes sense to have lambdas is the
  (7) infinitary function space, since we can only mention these in
  path types. But such lambdas can be always eliminated by introducing
  extra function parameters which are constrained by pointwise
  equalities. In the case of the mentioned example,

    z : H
    n : (A → H) → H
    q : n (λ x. z) = z

  We may have the following equivalent λ-free type for q:

    q : (f : A → H) → ((x : A) → f x = z) → n f = z

  In general, if we have
    Γ;Δ ⊢ λx. t : El ((x : a) → b)
  Then instead we may have
    Γ;Δ, f : El ((x : a) → b), p : (x : a) → f x = t ⊢ f : El ((x : a) → b)

  We include this explanation in the revision.

- Page 7, 8 item (5), about type equalities and UA

  We include this in the revision.

- Page 10, about initial NatAlg
  We use Nat* instead of initial Nat-algebra in the revised version.

- Page 11, about coherence issue, references

  In the revision we add a reference to the following discussion:

    https://homotopytypetheory.org/2014/03/03/hott-should-eat-itself/

  As discussed there, our coherence problem is related to the better
  known problem of constructing semisimplicial types; solving the
  former implies a solution of the latter.

- Page 12, about initial CwF

  We removed the problematic "initial" occurrence. To say a bit more
  here, if we set-truncate CwFs then there is a strict category of
  CwFs, and if we don't truncate then we conjecture that there is a
  wild category of wild CwF-s, but in any case we are referring to a
  1-category for the purpose of induction and recursion. We think that
  the subtleties involving these categories are not very relevant to
  the discussion in that section, so we avoid it in the revision.

- Page 13, "would merely decidable equality also be sufficient?"

  For the purpose of elaborating raw syntax into intrinsic syntax
  (which is in our view an important motivation for decidability), it
  would not be sufficient, because we need to eliminate from decided
  equalities to proof-relevant intrinsic terms.

--------------------------------------------------------------------------------

Response to the second reviewer:

- On the suggestion for identity homomorphisms:

  This suggestion does solve the U[] case. However, it still does not
  work; some cases (like U[]) become provable, where we can make use
  of strict equality hypotheses, but other cases fail, where we need
  to provide strict equalities. For example, assuming the suggested
  strict tᴵᴰ equations, we have

    (t =ₐ u)ᴵᴰ γ : (t = u)ᴹ γ γ (Γᴵᴰ γ) ≡ Uᴵᴰ γ ((t = u)ᴬ γ)

  The type can be reduced along definitions, and using tᴵᴰ and uᴵᴰ, to:

    (λ e. refl ◾ ap (λ x. x) e ◾ refl) ≡ (λ e. e)

  Here, e : tᴬ γ = uᴬ γ, and it is necessarily a non-strict equality. Hence,
  this case is not provable with strict tᴵᴰ equalities but provable with weak
  ones. We include this explanation in the revision.

- Addressing comments:
  - page 7, Figure 2, about external parameters:

    We do not construct HIIT definitions, only signatures. HIIT signatures can
    only depend on an HIIT specified by a previous signature if we assume the
    existence of that HIIT in the external type theory (i.e. as part of the
    context Γ). Also, the codomain types of external (6) and infinitary (7)
    functions can depend on the additional (x : A) external binder. For example,
    if we have Γ = (Nat: Type₀, zero: Nat, suc: Nat → Nat), and we write the
    signature for length-indexed A-vectors, then we have "cons : (n : Nat) → A →
    Vec n → Vec (suc n)", where the "suc n" expression refers back to the
    external "n : Nat" binder. The only way to extend the external Γ is by using
    (6) and (7) functions, otherwise, we can only refer to Γ entries which are
    assumed on the "top level", such as the A : Type₀ in the example for lists
    on page 8 in Section 2.1.

  - page 12: for the intrinsic syntax, do you also assume an induction
    principle for it, e.g. that it is initial?

    Yes. Of course, the induction principle for the intrinsic syntax would be
    something that is formalized in the current paper, so if we want to avoid
    circularity, then we need to say that the induction principle is given in an
    ad-hoc fashion. This is probably acceptable, since such ad-hoc HIT induction
    principles have been common in previous HoTT literature.

  - page 12, line -9: you say that you "need to" set-truncate; is this
    rather a strong belief that you have to?

    In the next paragraphs we justify the need for truncation. We
    think that undecidability of conversion and impossibility of
    extrinsic term models are solid justifications.

  - page 21, line -16: is there an expected equation if all computation rules
    were definitional?

    We think that if we follow "book" HoTT practice, p^s would be a
    propositional computation rule, because it involves "coe" which is defined
    externally, similarly to how path constructor computation rules involving
    "ap" are also propositional.

### 2019-12-02 paper.tex
typo discovered by Thorsten
revision 2 updates

\begin{abstract}
\noindent
Higher inductive-inductive types (HIITs) generalize inductive types of
dependent type theories in two ways. On the one hand they allow the
simultaneous definition of multiple sorts that can be indexed over
each other. On the other, hand they support equality constructors,
thus generalizing higher inductive types of homotopy type
theory. Examples that make use of both features are the Cauchy real
numbers and the well-typed syntax of type theory where conversion
rules are given as equality constructors. In this paper we propose a
general definition of HIITs using a small type theory, named the
theory of signatures. A context in this theory encodes a HIIT by
listing the constructors. We also compute notions of induction and
recursion for HIITs by using variants of syntactic logical relation
translations. Building full categorical semantics and constructing
initial algebras is left for future work. The theory of HIIT
signatures was formalised in Agda together with the syntactic
translations. We also provide a Haskell implementation, which takes
signatures as input and outputs translation results as valid Agda
code.
\end{abstract}

@@

(6) We have a function space where the domain is a type in the external
theory. We distinguish it from (3) by using red brick color in the domain
specification. We specify lists and integers as follows, with integers now
given as quotients of pairs of natural numbers:
\begin{alignat*}{4}
  & \hat{A}\in\Type_{\hat{0}} && \vdash\,\boldsymbol{\cdot},\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
  & \hat{\Gamma} && \vdash\,\boldsymbol{\cdot},\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
\end{alignat*}
In the case of integers, $\hat{\Gamma}$ is
$\hat{Nat}\in\Type_{\hat{0}}\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
numbers in the external theory. As another example, propositional
truncation for a type $\hat{A}$ is specified as follows.
\[
\hat{A}\in\Type_{\hat{0}}\vdash\, \boldsymbol{\cdot},\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
\]
The smallness of $\hat{A}$ is required in (6). It is possible to generalize
signatures to arbitrary universe levels, but it is not essential to the current
development. Note that the (6) function space preserves strict positivity, since
in the external theory there is no way to recursively refer to the inductive type
\emph{being defined}. The situation is analogous to the case of $W$-types
\cite{abbot05containers}, where shapes and positions can contain arbitrary types
but they cannot recursively refer to the $W$-type being defined.

@@

\subsubsection{Set-truncation}
Additionally, we need to set-truncate the syntax by adding the following constructors:
\begin{alignat*}{5}
  & \mathsf{setTy}  && : (A\,B : \Ty\,\Gamma)(p\,q : A = B)\ra p = q \\
  & \mathsf{setTm}  && : (t\,u : \Tm\,\Gamma\,A)(p\,q : t = u)\ra p = q \\
  & \mathsf{setSub} && : (\sigma\,\delta : \Sub\,\Gamma\,\Delta)(p\,q : \sigma = \delta)\ra p = q
\end{alignat*}

@@

We can easily show that the non-truncated syntax does not form sets. For
example, $\U[]$ and $[\id]$ are two proofs of $\U[\id] = \U$, and they are not
forced to be equal. Assuming univalence, we can give a model where types are
interpreted as closed metatheoretic types, $\U$ is interpreted as metatheoretic
$\Bool$, $[\id]$ is interpreted as $\refl : A = A$ for some metatheoretic $A$
type, and $\U[]$ is interpreted as the $\Bool$ negation equivalence, thereby
formally distinguishing $\U[]$ and $[\id]$. Hence, by Hedberg's theorem
\cite{hedberg}, the syntax does not have decidable equality. This also implies
that the non-truncated intrinsic syntax is not constructible from set quotients
of extrinsic terms (since those always form sets). The non-truncated syntax just
does not seem to be a sensible notion. This situation is similar to how
categories in homotopy type theory need to have set-truncated morphisms
\cite{ahrens2015univalent}.

@@

\[
\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
\]
The sphere-algebras are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\A \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
\end{alignat*}
Given a sphere-algebra and fixing a universe level $\hat{i}$, the displayed algebra is the following.
\begin{alignat*}{5}
  & && \Gamma^\D\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2D}}\in \hat{S^2}\rah\Type_{\hat{i}}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^D}\in \hat{S^{2D}\,b}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^D\in \TR_{\hat{(}\TR_{S^{2D}}\,\blank\,b^D=b^D\hat{)}}\,surf\,\refl_{b^D} = \refl_{b^D}}\hat{)}
\end{alignat*}
Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
\begin{alignat*}{5}
  & && \Gamma^\S\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2D},\,b^D,\,surf^D}\hat{)} \\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2S}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2D}\,x}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^S}\in \hat{S^{2S}\,\,b = b^D}\hat{)}\\
  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^S\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^S}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^S}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}\hat{.}\TR\,\hat{b^S}\,\hat{(}\TR\,\hat{b^S}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
  & && \hspace{5.5em} \hat{=} \,\hat{surf^D}\hat{\bigg)}
\end{alignat*}
Note that if $\hat{b^S}$ is a definitional equality (that is, we have
$\hat{S^{2S}\,\,b}\,\hat{\equiv}\,\hat{b^D}$), the occurrences of $\hat{b^S}$
in the type of $\hat{surf^S}$ can be replaced by $\hat{\refl}$. In
this case the type of $\hat{surf^S}$ becomes the more usual
$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^D}$.

>>>>

\begin{abstract}
\noindent
Higher inductive-inductive types (HIITs) generalize inductive types of
dependent type theories in two ways. On the one hand they allow the
simultaneous definition of multiple sorts that can be indexed over
each other. On the other hand they support equality constructors,
thus generalizing higher inductive types of homotopy type
theory. Examples that make use of both features are the Cauchy real
numbers and the well-typed syntax of type theory where conversion
rules are given as equality constructors. In this paper we propose a
general definition of HIITs using a small type theory, named the
theory of signatures. A context in this theory encodes a HIIT by
listing the constructors. We also compute notions of induction and
recursion for HIITs by using variants of syntactic logical relation
translations. Building full categorical semantics and constructing
initial algebras is left for future work. The theory of HIIT
signatures was formalised in Agda together with the syntactic
translations. We also provide a Haskell implementation, which takes
signatures as input and outputs translation results as valid Agda
code.
\end{abstract}

@@

(6) is a function space where the domain is a type in the external theory. We
distinguish it from (3) by using red brick color in the domain specification. We
specify lists and integers as follows, with integers now given as quotients of
pairs of natural numbers:
\begin{alignat*}{4}
  & \hat{A}\in\Type_{\hat{0}} && \vdash\,\boldsymbol{\cdot},\,\, && List:\U,\,\,\,nil:\underline{List},\,\,\,cons:(\hat{x}\in \hat{A})\ra List\ra\underline{List} \\
  & \hat{\Gamma} && \vdash\,\boldsymbol{\cdot},\,\,&& Int:\U,\,\,\,pair:(\hat{x}\,\hat{y}\in\hat{Nat})\ra\underline{Int},\,\,\, \\
  & && && eq:(\hat{a}\,\hat{b}\,\hat{c}\,\hat{d}\in \hat{Nat})(\hat{p}\in \hat{a}\hat{\mathbin{+}}\hat{d}\eqh_{\hat{Nat}} \hat{b}\hat{\mathbin{+}}\hat{c})\ra \underline{pair\,\hat{a}\,\hat{b}=_{Int}\mathsf{pair}\,\hat{c}\,\hat{d}}, \\
  & && && trunc:(x y : Int)(p\,q : a=_{Int} b)\ra \underline{p=_{x=_{Int} y} q}
\end{alignat*}
In the case of integers, $\hat{\Gamma}$ is
$\hat{Nat}\in\Type_{\hat{0}}\hat{,}\,\blank\hat{\mathbin{+}}\blank\in\hat{Nat}\rah
\hat{Nat}\rah\hat{Nat}$, or alternatively, we could require natural
numbers in the external theory. As another example, propositional
truncation for a type $\hat{A}$ is specified as follows.
\[
\hat{A}\in\Type_{\hat{0}}\vdash\, \boldsymbol{\cdot},\,\,\,tr:\U,\,\,\,emb : (\hat{x}\in \hat{A})\ra \underline{tr},\,\,\,eq:(x\,y:tr)\ra \underline{x=_{tr} y}
\]
The smallness of $\hat{A}$ is required in (6). It is possible to generalize
signatures to arbitrary universe levels, but it is not essential to the current
development.

Note that we can assume arbitrary structures in the external $\hat{\Gamma}$
context, which in particular allows us to specify HIITs depending on other
(external) HIITs. We can do this by first specifying a HIIT, then using the
methods in this paper to compute notions of algebras and induction, then assume
the HIIT in the external context of another HIIT signature.

The (6) function space preserves strict positivity, since in the external theory
there is no way to recursively refer to the inductive type \emph{being
  defined}. The situation is analogous to the case of $W$-types
\cite{abbot05containers}, where shapes and positions can contain arbitrary types
but they cannot recursively refer to the $W$-type being defined. This setup rules
out some signatures; for example, rose trees cannot be specified as follows, because
we cannot apply the external $\hat{List}$ to the inductive $T$:
\[
\hat{A : \Type_0,\,\,List : \Type_0 \ra \Type_0} \vdash
     \boldsymbol{\cdot},\,\,T : \U,\,\,node : \hat{A} \ra \hat{List}\,T \ra \underline{T}
\]
Analogously nested HIT examples are the ``hubs and spokes'' definitions in
\cite[Section 6.7]{HoTTbook}. To allow such definitions, we would have to
analyze external constructions to check whether they preserve strict
positivity. This is out of the scope of the current paper.

@@

\subsubsection{Set-truncation}
Additionally, we prefer to set-truncate the syntax by adding the following
constructors, for reasons shortly explained:
\begin{alignat*}{5}
  & \mathsf{setTy}  && : (A\,B : \Ty\,\Gamma)(p\,q : A = B)\ra p = q \\
  & \mathsf{setTm}  && : (t\,u : \Tm\,\Gamma\,A)(p\,q : t = u)\ra p = q \\
  & \mathsf{setSub} && : (\sigma\,\delta : \Sub\,\Gamma\,\Delta)(p\,q : \sigma = \delta)\ra p = q
\end{alignat*}

@@

We can easily show that the non-truncated syntax does not form sets. For
example, $\U[]$ and $[\id]$ are two proofs of $\U[\id] = \U$, and they are not
forced to be equal. Assuming univalence, we can give a model where types are
interpreted as closed metatheoretic types, $\U$ is interpreted as metatheoretic
$\Bool$, $[\id]$ is interpreted as $\refl : A = A$ for some metatheoretic $A$
type, and $\U[]$ is interpreted as the $\Bool$ negation equivalence, thereby
formally distinguishing $\U[]$ and $[\id]$. Hence, by Hedberg's theorem
\cite{hedberg}, the syntax does not have decidable equality, and hence it does
not support decidable type checking. This also implies that the non-truncated
intrinsic syntax is not constructible from set quotients of extrinsic terms
(since those always form sets). The non-truncated syntax just does not seem to
be a sensible notion. This situation is similar to how categories in homotopy
type theory need to have set-truncated morphisms \cite{ahrens2015univalent}.

@@

\[
\Gamma :\equiv \Big(\boldsymbol{\cdot},\,\,\,S^2:\U,\,\,\,b:\underline{S^2},\,\,\,surf:\underline{\refl_{b} =_{(b=_{S^2} b)} \refl_{b}}\Big)
\]
The sphere-algebras are computed as follows.
\begin{alignat*}{5}
  & && \Gamma^\A \equiv \hat{\top}\timesh\hat{(}\hat{S^2}\in\Type_{\hat{0}}\hat{)}\timesh\hat{(}\hat{b}\in \hat{S^2}\hat{)}\timesh\hat{(}\hat{surf\in \hat{\refl_{\hat{b}}}=_{\hat{(}\hat{b}=_{\hat{S^2}}\hat{b}\hat{)}}\hat{\refl_{\hat{b}}}}\hat{)}
\end{alignat*}
Given a sphere-algebra and fixing a universe level $\hat{i}$, the displayed algebra is the following.
\begin{alignat*}{5}
  & && \Gamma^\D\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2D}}\in \hat{S^2}\rah\Type_{\hat{i}}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^D}\in \hat{S^{2D}\,b}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{surf^D\in \TR_{\hat{(}\TR_{S^{2D}}\,\blank\,b^D=b^D\hat{)}}\,surf\,\refl_{b^D} = \refl_{b^D}}\hat{)}
\end{alignat*}
Given a sphere-algebra and a displayed algebra over it, we get the type of sections:
\begin{alignat*}{5}
  & && \Gamma^\S\,\,\hat{(}\hat{\tt,\,S^2,\,b,\,surf}\hat{)}\,\hat{(}\hat{\tt,\,S^{2D},\,b^D,\,surf^D}\hat{)} \\
  & \equiv \,\, && \hat{\top}\timesh\hat{(}\hat{S^{2S}}\in \hat{(}\hat{x}\in\hat{S^2}\hat{)}\hat{\,\ra\,}\hat{S^{2D}\,x}\hat{)} \\
  & && \hspace{0.75em} \timesh\hat{(}\hat{b^S}\in \hat{S^{2S}\,\,b = b^D}\hat{)}\\
  & && \hspace{0.75em} \timesh\hat{\bigg(}\hat{surf^S\,\,\in\,\,} \TR\,\hat{\Big(}\hat{\J\,\refl\,b^S}\hat{\Big)}\,\hat{\Big(}\TR\,\hat{\big(}\hat{\J\,\refl\,b^S}\hat{\big)}\,\hat{\big(}\hat{\apd}\,\hat{(}\hat{\lambda} \hat{x}\hat{.}\TR\,\hat{b^S}\,\hat{(}\TR\,\hat{b^S}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\,\hat{x}\hat{)}\hat{)}\hat{)}\,\hat{surf}\hat{\big)}\hat{\Big)} \\
  & && \hspace{5.5em} \hat{=} \,\hat{surf^D}\hat{\bigg)}
\end{alignat*}
Note that if $\hat{b^S}$ is a definitional equality (that is, we have
$\hat{S^{2S}\,\,b}\,\hat{\,\equiv\,}\,\hat{b^D}$), the occurrences of $\hat{b^S}$
in the type of $\hat{surf^S}$ can be replaced by $\hat{\refl}$. In
this case the type of $\hat{surf^S}$ becomes the more usual
$\hat{\apd}\,\hat{(}\hat{\apd}\,\hat{S^{2S}}\hat{)}\,\hat{surf}\,\hat{=}\,\hat{surf^D}$.
